Anxiety, avoidance, and sequential evaluation

### 2019-04-14 manuscripts/manuscript/manuscript.tex

\documentclass[11pt]{article} % For LaTeX2e
\usepackage{rldm2019, palatino}
\usepackage{graphicx}
\usepackage{amsfonts, amsmath}
\usepackage{algorithm, algpseudocode}%
\usepackage[numbers,super]{natbib}

@@

@@

Of note in this respect, anxiety and depression are highly comorbid, with roughly 45\% of individuals with a lifetime depression diagnosis also diagnosed with an anxiety disorder \citep{kessler2015}. One interesting proposal is that this association often arises longitudinally: in particular, that anxiety precedes certain types of depression \citep{alloy1990, jacobson2014}. The idea, in brief, is that anxiety begets avoidance and foregone reward, leading ultimately to a belief that reward is unattainable and subsequently depression. This story is borne out by simulations of learning in our model (not shown, for space) in environments like that of Fig.~1: the penumbra of negative value under pessimistic assumptions spreads gradually throughout the environment with the progress of learning, ultimately leading the agent to expect no reward and, echoing anergic symptoms of depression, forego action altogether.

@@

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/04_tree_rldm.png}}%
  }
  \par \textbf{Figure 3:} (A) The decision tree environment from \cite{Huys2012} and \cite{Lally2017}. An optimistic agent ($w=1$) prefers the optimal loss-minimizing policy through the initial large loss (left branch). (B) A pessimistic agent ($w=0.5$) comes to prefer the branch without the large loss so as to avoid being unable to recoup the large initial loss. One-step rewards are presented in each state; the net value $Q$ on each path is shown numerically. (Parameters: $\gamma = 1.0$)
\end{figure}

@@

@@

\bibliographystyle{naturemag}
\small{\bibliography{rldm2019}}

>>>>

\documentclass[11pt]{article} % For LaTeX2e
\usepackage{manuscript, palatino}
\usepackage{graphicx}
\usepackage{amsfonts, amsmath}
\usepackage{algorithm, algpseudocode}%
\usepackage[numbers,super]{natbib}

@@

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/05_tree_rldm.png}}%
  }
  \par \textbf{Figure 3:} (A) The decision tree environment from \cite{Huys2012} and \cite{Lally2017}. An optimistic agent ($w=1$) prefers the optimal loss-minimizing policy through the initial large loss (left branch). (B) A pessimistic agent ($w=0.5$) comes to prefer the branch without the large loss so as to avoid being unable to recoup the large initial loss. One-step rewards are presented in each state; the net value $Q$ on each path is shown numerically. (Parameters: $\gamma = 1.0$)
\end{figure}

@@

Of note in this respect, anxiety and depression are highly comorbid, with roughly 45\% of individuals with a lifetime depression diagnosis also diagnosed with an anxiety disorder \citep{kessler2015}. One interesting proposal is that this association often arises longitudinally: in particular, that anxiety precedes certain types of depression \citep{alloy1990, jacobson2014}. The idea, in brief, is that anxiety begets avoidance and foregone reward, leading ultimately to a belief that reward is unattainable and subsequently depression. This story is borne out by simulations of learning in our model (Fig.~4) in environments like that of Fig.~1: the penumbra of negative value under pessimistic assumptions spreads gradually throughout the environment with the progress of learning, ultimately leading the agent to expect no reward and, echoing anergic symptoms of depression, forego action altogether.

@@

\begin{figure}[!b]
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../../figures/04_lh.png}}%
  }
  \par \textbf{Figure 3:} (A) The decision tree environment from \cite{Huys2012} and \cite{Lally2017}. An optimistic agent ($w=1$) prefers the optimal loss-minimizing policy through the initial large loss (left branch). (B) A pessimistic agent ($w=0.5$) comes to prefer the branch without the large loss so as to avoid being unable to recoup the large initial loss. One-step rewards are presented in each state; the net value $Q$ on each path is shown numerically. (Parameters: $\gamma = 1.0$)
\end{figure}

@@

The two accounts described above are by no means the only possible reinforcement learning formalisms that can give rise to in threat-avoidance behavior. Traditionally, reinforcement learning algorithms aim to find reward-maximizing policies with respect to the expected (average) returns. However, in environments with the potential for catastrophic loss it may be preferable to optimize a policy with respect to an alternative objective function insofar that the expected return does not necessarily avoid rare yet large negative outcomes. Such risk-sensitive reinforcement learning algorithms have been the subject of much recent research \cite{geibel2005, morimura2012} and show threat-avoidance behaviors not unlike those presented here. An often reported clinical observation is that anxious individuals disproportionately dwell upon low chance yet high cost possibilities. \cite{Miceli2005} One speculative hypothesis is that anxious individuals optimize with respect to a more conservative risk criterion.

A related possibility concerns the notion of state itself. Here we assumed that states were fully observable, such that there was no uncertainty regarding the state an agent was currently occupying. In real-world environments, however, the state must be inferred from external and internal variables. As has been proposed elsewhere \cite{Paulus2012}, one possibility is that anxiety biases state inference resulting in erroneous beliefs about likely future events. There is evidence that anxious individuals are biased in their sampling and interpretation of incoming threat information \cite{Paulus2012}, suggesting that the avoidance behaviors may be a function of pessimistic beliefs about the current state of the environment.

Of course, none of the above constitutes a claim about the biological mechanisms underlying pessimistic sequential evaluation. One possibility is that these symptoms reflect biased model-based planning. Pathological worry is strongly associated with anxiety disorders and may be viewed as a biased form of future action planning. In line with the present results, higher levels of worry is associated with reduced self-efficacy beliefs, greater endorsement of statements concerning personal inadequacies in the face of threat, and that the association between perceived competence and worry is mediated by exaggerated threat appraisal (for review, see \cite{Berenbaum2010}). As such, the behaviors observed above may reflect biased sampling of possible future outcomes as a function of pessimistic beliefs about self-efficacy.

As alluded to above, a second possibility is that biased sequential evaluation may arise as a function of biased offline replay of events. During periods of rest and sleep, the hippocampus replays recent experiences. Recent modeling has shown that hippocampal replay is in part driven by gain from propagating newly encountered information to preceding locations. Biases in the tagging of memories for replay, as a function of either pessimistic expected future state occupancy or biased learning of information, may yield the propagation of threat value similarly as has been observed above.

@@

\bibliographystyle{naturemag}
\small{\bibliography{manuscript}}
